---
title: "Acoustic Feature Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Acoustic Feature Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

This vignette covers three ASAP functions for examining acoustic structure within
a single recording: `spectral_entropy()`, `FF()`, and `amp_env()`.

- **`spectral_entropy()`** measures how tonal or noisy a sound is at each point
  in time by quantifying how evenly power is spread across frequencies.
  Zebra finch syllables are often highly structured and may show low entropy, 
  especially for harmonic stack syllables or certain calls; in contrast, noisier 
  syllables, broadband calls, or background noise tend to produce higher-entropy 
  traces.
- **`FF()`** tracks fundamental frequency (perceived pitch) over time. It is the
  go-to tool for looking at pitch contours within a single syllable or comparing
  pitch trajectories across many renditions.
- **`amp_env()`** extracts the amplitude envelope, summarizing how the overall
  intensity of sound rises and falls across the duration of a segment.

The vignette walks through how to call each function on a simple time window
(provide a file path and start/end times), and then shows how the same functions
accept a segment data frame produced by `segment()` — a natural bridge toward
population-level analysis with SAP objects.

**Prerequisites**: Before reading this vignette, we recommend completing:

- [Overview: ASAP 101](single_wav_analysis.html) - Basic ASAP functions

**What you will learn**:

1. What kinds of data each function accepts
2. How to measure spectral structure with entropy (Wiener and Shannon)
3. How to extract pitch contours with the cepstrum method
4. How to measure amplitude envelopes and reuse segment data frames for
   feature analysis

---

## Overview: What Data Do These Functions Accept?

All three functions share a common design: they dispatch on the class of their
first argument, so you can pass different kinds of input without changing the
function name.

### `spectral_entropy()` and `FF()`

| Input type | What to pass | Key extra argument |
|---|---|---|
| Single WAV file | Character path to `.wav` file | `start_time`, `end_time` (seconds) |
| Segment data frame | Data frame with `filename`, `start_time`, `end_time` columns | `wav_dir` (directory containing the WAV files) |
| SAP object | A `Sap` object | `segment_type`, sampling/filtering options |
| Pre-computed matrix | An entropy or F0 matrix returned by a previous call | — (re-plots the stored matrix) |

When `x` is a **character path**, `start_time` and `end_time` default to the
full file duration if omitted.

When `x` is a **data frame**, it must contain at least `filename`, `start_time`,
and `end_time` columns. A single-row data frame behaves identically to the WAV
file method. A multi-row data frame triggers parallel processing and returns an
aligned matrix.

### `amp_env()`

`amp_env()` always takes a **single-row data frame** (`segment_row`) — the row
must contain at least `start_time` and `end_time`. Provide `wav_dir` when the
WAV file directory is not embedded in the data frame as an attribute.

```
amp_env(segment_row, wav_dir = NULL, msmooth = NULL,
        amp_normalize = c("none", "peak", "rms"), plot = FALSE)
```

---

## Setup

```{r setup}
library(ASAP)

wav_file <- system.file("extdata", "zf_example.wav", package = "ASAP")
analysis_start <- 1.5
analysis_end <- 2.5
```

---

## 1. Spectral Entropy Analysis

Spectral entropy measures how structured or noisy the frequency distribution is
within a sound segment. Harmonic syllables tend to have lower entropy, while
broader-band noisy sounds tend to have higher entropy.

### Choosing `method` and `normalize`

Two arguments control what the trace looks like:

- **`method`**: `"wiener"` (default) quantifies spectral flatness — more negative
  values mean more structured sound, 0 means noise-like. `"shannon"` quantifies
  information content — higher values reflect more uniform spectral energy.
- **`normalize`**: `FALSE` returns the native scale; `TRUE` rescales the output
  to 0–1, making it easier to compare plots across different recordings or methods.

These arguments are independent — you can use either method with or without
normalization:

```{r spectral-entropy-wiener-raw, fig.height=5}
# Wiener entropy, native scale
wiener_raw <- spectral_entropy(
  wav_file,
  start_time = analysis_start,
  end_time   = analysis_end,
  method     = "wiener",
  normalize  = FALSE,
  plot       = TRUE
)
```

```{r spectral-entropy-wiener-norm, fig.height=5}
# Wiener entropy, normalized to 0–1
wiener_norm <- spectral_entropy(
  wav_file,
  start_time = analysis_start,
  end_time   = analysis_end,
  method     = "wiener",
  normalize  = TRUE,
  plot       = TRUE
)
```

```{r spectral-entropy-shannon-raw, fig.height=5}
# Shannon entropy, native scale
shannon_raw <- spectral_entropy(
  wav_file,
  start_time = analysis_start,
  end_time   = analysis_end,
  method     = "shannon",
  normalize  = FALSE,
  plot       = TRUE
)
```

```{r spectral-entropy-shannon-norm, fig.height=5}
# Shannon entropy, normalized to 0–1
shannon_norm <- spectral_entropy(
  wav_file,
  start_time = analysis_start,
  end_time   = analysis_end,
  method     = "shannon",
  normalize  = TRUE,
  plot       = TRUE
)
```

### Argument guide

| Argument | Options | How to think about it |
|---|---|---|
| `method` | `"wiener"` or `"shannon"` | Use Wiener for spectral flatness (classic bioacoustics metric); use Shannon for information-content style entropy |
| `normalize` | `FALSE` or `TRUE` | Use `FALSE` to keep the native scale; use `TRUE` for a 0–1 scale that is easier to compare across plots |

All four combinations of `method` × `normalize` are valid; pick whichever suits
your analysis question.

### Quality check

Use the plot to compare low-entropy tonal structure against higher-entropy
noisy regions. If the segment includes long silent stretches, tighten the time
window so the estimate reflects the vocalization itself rather than surrounding
silence.

---

## 2. Fundamental Frequency (Pitch) Analysis

The `FF()` function extracts the fundamental frequency contour, showing how
perceived pitch changes over time.

The two most important arguments for a single recording are:

- `method`: `"cepstrum"` or `"yin"`
- `fmax`: the upper pitch limit to search (Hz)

### Cepstrum method

`"cepstrum"` is the default and the easiest place to start. It is fast and
works well for quick inspection of tonal syllables.

```{r ff-cepstrum, fig.height=5}
pitch_cepstrum <- FF(
  wav_file,
  start_time = analysis_start,
  end_time   = analysis_end,
  method     = "cepstrum",
  fmax       = 1400,
  threshold  = 10,
  plot       = TRUE
)
```

### YIN method

`"yin"` can be more robust for some signals, but it requires Python
dependencies through `reticulate`, including `librosa` and `numpy`.

```{r ff-yin, fig.height=5}
# ASAP attempts to auto-install librosa/numpy via reticulate when needed.
# This chunk runs only when Python and its dependencies are available;
# it skips gracefully otherwise (e.g. on CRAN or CI build servers).
has_yin <- tryCatch({
  requireNamespace("reticulate", quietly = TRUE) &&
    reticulate::py_module_available("librosa") &&
    reticulate::py_module_available("numpy")
}, error = function(e) FALSE)

if (has_yin) {
  pitch_yin <- FF(
    wav_file,
    start_time = analysis_start,
    end_time   = analysis_end,
    method     = "yin",
    fmax       = 1400,
    threshold  = 10,
    plot       = TRUE
  )
} else {
  message("YIN method requires Python with librosa and numpy. Skipping.")
}
```

### Argument guide

| Argument | Options | How to think about it |
|---|---|---|
| `method` | `"cepstrum"` or `"yin"` | Start with `cepstrum`; try `yin` if you want an alternative pitch tracker and have Python dependencies installed |
| `fmax` | Numeric upper limit in Hz | Raise it if the contour clips too low; lower it to reduce implausibly high estimates |
| `threshold` | Confidence filter (%) | Higher values remove uncertain estimates but may introduce gaps |

### The result contains

- `f0`: Fundamental frequency values over time (kHz)
- `time`: Corresponding time stamps (seconds)

---

## 3. Amplitude Envelope from a Syllable Data Frame

The amplitude envelope summarizes how sound intensity changes over time.
`amp_env()` takes a single-row segment data frame with `filename`, `start_time`,
and `end_time` columns. A syllable data frame from `segment()` is a simple way
to create that input.

### Step 1: Segment a small time window into syllables

```{r amplitude-envelope-segment}
syllables <- segment(
  wav_file,
  start_time         = 1,
  end_time           = 5,
  flim               = c(1, 8),
  silence_threshold  = 0.01,
  min_syllable_ms    = 20,
  max_syllable_ms    = 240,
  min_level_db       = 10,
  verbose            = FALSE,
  plot               = FALSE
)

knitr::kable(head(syllables), digits = 3)
```

### Step 2: Choose one syllable row

```{r amplitude-envelope-row}
example_syllable <- NULL
if (!is.null(syllables) && nrow(syllables) >= 1) {
  example_syllable <- syllables[1, , drop = FALSE]
  knitr::kable(example_syllable, digits = 3)
}
```

### Step 3: Extract the envelope

```{r amplitude-envelope-syllable, fig.height=4}
if (!is.null(example_syllable)) {
  env_syl <- amp_env(
    example_syllable,
    wav_dir       = dirname(wav_file),
    msmooth       = c(256, 50),
    amp_normalize = "peak",
    plot          = TRUE
  )
}
```

### Smoothing and normalization

- **`msmooth`**: a length-2 vector `c(window_samples, overlap_percent)`. Larger
  windows smooth out fine-grained amplitude fluctuations; smaller windows
  preserve rapid transients.
- **`amp_normalize`**: `"none"` keeps the raw scale; `"peak"` scales to the
  maximum amplitude (good for comparing envelope shapes); `"rms"` scales by
  RMS energy (good for comparing absolute loudness).

### Argument guide

| Argument | Options | How to think about it |
|---|---|---|
| `segment_row` | Single-row data frame | Must contain `filename`, `start_time`, `end_time` |
| `wav_dir` | Path string | Directory containing the WAV files |
| `msmooth` | Numeric vector c(window, overlap) | Larger window = smoother envelope; smaller = more temporal detail |
| `amp_normalize` | `"none"`, `"peak"`, `"rms"` | Use `"peak"` when comparing envelope shapes across segments |

---

## 4. Add-on: Entropy and Pitch from a Segment Data Frame

The same syllable table can also be passed directly to `spectral_entropy()` and
`FF()`. This is useful when you want to analyze a set of detected segments
instead of manually specifying `start_time` and `end_time` for each one.

### Step 1: Select a few syllable rows

```{r feature-data-frame-rows}
example_syllables <- NULL
if (!is.null(syllables) && nrow(syllables) >= 3) {
  example_syllables <- syllables[1:3, ]
  knitr::kable(example_syllables, digits = 3)
}
```

### Step 2: Run spectral entropy on the data frame

The same `method` and `normalize` arguments from the single-file examples
apply here. Mix and match freely — for example, try `method = "shannon"` with
`normalize = FALSE` if you prefer the information-content scale.

```{r spectral-entropy-data-frame, fig.height=5}
if (!is.null(example_syllables)) {
  entropy_df <- spectral_entropy(
    example_syllables,
    wav_dir   = dirname(wav_file),
    method    = "wiener",
    normalize = TRUE,
    plot      = TRUE
  )
}
```

### Step 3: Run pitch analysis on the data frame

For data-frame input, `FF()` aligns the selected segments onto a common time
axis and returns a multi-segment result. The `method`, `fmax`, and `threshold`
arguments work the same way as for a single WAV file.

```{r ff-data-frame, fig.height=5}
if (!is.null(example_syllables)) {
  pitch_df <- FF(
    example_syllables,
    wav_dir   = dirname(wav_file),
    method    = "cepstrum",
    fmax      = 1400,
    threshold = 10,
    plot      = TRUE
  )
}
```

---

## Summary

These three functions provide a compact toolkit for exploring acoustic
variation in a single recording:

| Function | What it captures | Primary input types |
|---|---|---|
| `spectral_entropy()` | Spectral structure or noisiness | WAV path, data frame, SAP object, pre-computed matrix |
| `FF()` | Pitch contour over time | WAV path, data frame, SAP object, pre-computed matrix |
| `amp_env()` | Amplitude dynamics within a segment | Single-row data frame |

For motif-scale acoustic analysis across many recordings, continue to the SAP
object workflow starting with [Constructing SAP Object](construct_sap_object.html).

## Session Info

```{r session-info}
sessionInfo()
```
