---
title: "Acoustic Feature Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Acoustic Feature Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

This vignette focuses on acoustic measurements you can extract from a single
recording after identifying bouts or syllables of interest. It pairs naturally
with [Overview: ASAP 101](single_wav_analysis.html) and can also be used after
[Motif Detection](motif_detection.html) when you want to inspect the structure
of selected song segments in more detail.

**Prerequisites**: Before reading this vignette, we recommend completing:

- [Overview: ASAP 101](single_wav_analysis.html) - Basic ASAP functions

**What you will learn**:

1. How to measure spectral structure with entropy
2. How to extract pitch contours from a single song segment
3. How to compare amplitude-envelope shapes across bouts and syllables

---

## Overview

These feature-analysis functions are best used after you have already located a
time window, bout, or syllable worth inspecting. Because this is a single-WAV
tutorial, the vignette stays focused on a few small runnable examples rather
than wrapping everything in a larger umbrella workflow. The aim is to show what
each feature captures and how to interpret the resulting plots.

---

## Setup

```{r setup}
library(ASAP)

wav_file <- system.file("extdata", "zf_example.wav", package = "ASAP")

bouts <- find_bout(
  wav_file,
  rms_threshold = 0.1,
  min_duration = 0.7,
  plot = FALSE
)

syllables <- segment(
  wav_file,
  start_time = 1,
  end_time = 5,
  flim = c(1, 8),
  silence_threshold = 0.01,
  min_syllable_ms = 20,
  max_syllable_ms = 240,
  min_level_db = 10,
  verbose = FALSE,
  plot = FALSE
)
```

---

## 1. Spectral Entropy Analysis

Spectral entropy measures how structured or noisy the frequency distribution is
within a sound segment. Harmonic syllables tend to have lower entropy, while
broader-band noisy sounds tend to have higher entropy.

```{r spectral-entropy, fig.height=5}
entropy_result <- spectral_entropy(
  wav_file,
  start_time = 1.5,
  end_time = 2.5,
  method = "wiener",
  normalize = TRUE,
  plot = TRUE
)
```

### Interpreting entropy values

- **Low entropy (near 0)**: Structured, harmonic sounds
- **High entropy (near 1)**: Noisy, unstructured sounds

### Available methods

- `"wiener"`: Ratio of geometric to arithmetic mean
- `"shannon"`: Information-theoretic entropy

### Quality check

Use the plot to compare low-entropy tonal structure against higher-entropy
noisy regions. If the segment includes long silent stretches, tighten the time
window so the estimate reflects the vocalization itself rather than surrounding
silence.

---

## 2. Fundamental Frequency (Pitch) Analysis

The `FF()` function extracts the fundamental frequency contour, which describes
how perceived pitch changes over time.

```{r fundamental-frequency, fig.height=5}
pitch_result <- FF(
  wav_file,
  start_time = 1.5,
  end_time = 2.5,
  method = "cepstrum",
  fmax = 1400,
  threshold = 10,
  plot = TRUE
)
```

### Key parameters

- `fmax`: Maximum fundamental frequency to detect
- `threshold`: Higher values filter out uncertain estimates
- `method`: `"cepstrum"` (default) or `"yin"`

### The result contains

- `f0`: Fundamental frequency values over time
- `time`: Corresponding time stamps

### Tuning tips

| Parameter | Role | Typical effect |
|-----------|------|----------------|
| `fmax` | Upper bound for pitch search | Raise it if the contour clips at the top; lower it to avoid implausibly high estimates |
| `threshold` | Confidence filter | Higher values remove uncertain estimates but may leave gaps |
| `method` | Pitch extraction algorithm | `"cepstrum"` is a good default for quick inspection |

---

## 3. Amplitude Envelope

The amplitude envelope summarizes how sound intensity changes over time. This is
useful for comparing bout or syllable shapes after you have segmented them.

### Bout-level envelope

```{r amplitude-envelope-bout, fig.height=4}
if (!is.null(bouts) && nrow(bouts) >= 1) {
  env_bout <- amp_env(
    bouts[1, ],
    wav_dir = dirname(wav_file),
    msmooth = c(256, 50),
    amp_normalize = "peak",
    plot = TRUE
  )
}
```

### Syllable-level envelope

```{r amplitude-envelope-syllable, fig.height=4}
if (!is.null(syllables) && nrow(syllables) >= 1) {
  env_syl <- amp_env(
    syllables[1, ],
    wav_dir = dirname(wav_file),
    msmooth = c(256, 50),
    amp_normalize = "peak",
    plot = TRUE
  )
}
```

### Smoothing parameters

The `msmooth` argument controls envelope smoothing:

- First value: Window length in samples
- Second value: Overlap percentage

Smoother envelopes are useful for broad temporal structure, while lighter
smoothing preserves fine timing differences between syllables.

---

## Summary

These three functions provide a compact toolkit for exploring acoustic
variation in a single recording:

| Function | What it captures |
|----------|------------------|
| `spectral_entropy()` | Spectral structure or noisiness |
| `FF()` | Pitch contour over time |
| `amp_env()` | Amplitude dynamics within bouts or syllables |

For motif-scale acoustic analysis across many recordings, continue to the SAP
object workflow starting with [Constructing SAP Object](construct_sap_object.html).

## Session Info

```{r session-info}
sessionInfo()
```
