---
title: "Advanced Acoustic Feature Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Acoustic Feature Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

This vignette demonstrates an end-to-end workflow for extracting quantitative
acoustic features from a set of detected song motifs. The pipeline has two
phases:

1. **Motif detection** — run template-based detection on a single WAV file,
   define motif boundaries with `find_motif()`, and identify a syllable of
   interest within the motif.
2. **Feature extraction** — iterate over every detected motif and measure
   spectral entropy, fundamental frequency, and amplitude envelope within the
   selected syllable window; then summarize and plot how those features change
   across consecutive renditions.

The emphasis throughout is on structuring the output as tidy data frames that
are immediately ready for downstream statistical analysis.

**Prerequisites**: Before reading this vignette, we recommend completing:

- [Overview: ASAP 101](single_wav_analysis.html)
- [Motif Detection](motif_detection.html)
- [Acoustic Feature Analysis](acoustic_feature_analysis.html)

**What you will learn**:

1. How to recap motif detection in a few key lines
2. How to define a syllable window of interest relative to motif boundaries
3. How to iterate `spectral_entropy()`, `FF()`, and `amp_env()` over all
   motifs to build a per-motif feature table
4. How to summarize and visualize feature trajectories across renditions

---

## Setup

```{r setup}
library(ASAP)

wav_file <- system.file("extdata", "zf_example.wav", package = "ASAP")
wav_dir  <- dirname(wav_file)
```

---

## Phase 1: Motif Detection

The visualization and parameter-tuning steps are covered in detail in the
[Motif Detection](motif_detection.html) vignette. Here we recap only the
core lines needed to produce the `motifs` table that drives Phase 2.

```{r motif-detection}
# 1. Extract a reference clip and build a template from syllable "d"
clip_path <- create_audio_clip(wav_file, start_time = 1, end_time = 2.5)

template <- create_template(clip_path,
  template_name  = "d",
  start_time     = 0.72,
  end_time       = 0.84,
  freq_min       = 1,
  freq_max       = 10,
  write_template = FALSE
)

# 2. Detect template occurrences across the full recording
template_matches <- detect_template(
  x                = wav_file,
  template         = template,
  proximity_window = 1,
  save_plot        = FALSE
)

# 3. Expand each detection into full-motif boundaries
motifs <- find_motif(template_matches,
  pre_time = 0.7,
  lag_time = 0.5,
  wav_dir  = wav_dir
)

knitr::kable(motifs, digits = 3)
```

---

## Phase 2: Selecting a Syllable Window of Interest

Rather than analyzing the entire motif, we focus on a specific syllable to
reduce noise and make comparisons more precise. We define the syllable window
as a fixed offset **relative to each motif's `start_time`**.

Based on visual inspection of the motif spectrogram (see the
[Motif Detection](motif_detection.html) vignette), syllable "e" falls
roughly 0.78–0.85 s after the motif onset. We encode these offsets once so
they are applied consistently across every motif row.

```{r syllable-window}
# Offsets from each motif's start_time (seconds) — syllable "e"
syl_pre  <- 0.78   # syllable onset relative to motif start
syl_post <- 0.85   # syllable offset relative to motif start

# Build a per-motif data frame for the syllable window
syl_windows <- motifs
syl_windows$start_time <- motifs$start_time + syl_pre
syl_windows$end_time   <- motifs$start_time + syl_post
syl_windows$duration   <- syl_post - syl_pre

knitr::kable(syl_windows[, c("filename", "start_time", "end_time", "duration")],
             digits = 3)
```

Visualize the first motif's syllable "e" window to confirm the offset is
correctly aligned before applying it to all renditions:

```{r visualize-syllable-window, fig.height=4, fig.width=4, out.width="45%", fig.align="center"}
visualize_song(wav_file,
  start_time_in_second = syl_windows$start_time[1],
  end_time_in_second   = syl_windows$end_time[1]
)
```

## Phase 3: Feature Extraction Across All Motifs

We now iterate over every row of `syl_windows` and compute three features.
Each function accepts a single-row data frame plus `wav_dir`, so the pattern
is uniform for all three.

### 3a. Spectral Entropy

```{r entropy-loop}
entropy_results <- vector("list", nrow(syl_windows))

for (i in seq_len(nrow(syl_windows))) {
  row <- syl_windows[i, , drop = FALSE]

  # spectral_entropy() returns a 2-column matrix: [time, entropy]
  ent_mat <- spectral_entropy(
    row,
    wav_dir    = wav_dir,
    method     = "wiener",
    normalize  = FALSE,
    plot       = FALSE
  )

  entropy_results[[i]] <- data.frame(
    motif_index  = i,
    start_time   = row$start_time,
    mean_entropy = mean(ent_mat[, 2], na.rm = TRUE),
    sd_entropy   = sd(ent_mat[, 2],   na.rm = TRUE)
  )
}

entropy_df <- do.call(rbind, entropy_results)
knitr::kable(entropy_df, digits = 4)
```

### 3b. Fundamental Frequency

```{r ff-loop}
ff_results <- vector("list", nrow(syl_windows))

for (i in seq_len(nrow(syl_windows))) {
  row <- syl_windows[i, , drop = FALSE]

  # FF() returns a 2-column matrix: [time, frequency (kHz)]
  ff_mat <- FF(
    row,
    wav_dir   = wav_dir,
    method    = "cepstrum",
    fmax      = 1400,
    threshold = 10,
    plot      = FALSE
  )

  # Exclude zeros (unvoiced frames)
  voiced <- ff_mat[ff_mat[, 2] > 0, 2]

  ff_results[[i]] <- data.frame(
    motif_index = i,
    start_time  = row$start_time,
    mean_ff_kHz = ifelse(length(voiced) > 0, mean(voiced, na.rm = TRUE), NA_real_),
    sd_ff_kHz   = ifelse(length(voiced) > 0, sd(voiced,   na.rm = TRUE), NA_real_)
  )
}

ff_df <- do.call(rbind, ff_results)
knitr::kable(ff_df, digits = 4)
```

### 3c. Amplitude Envelope

We summarize the envelope by its peak amplitude and the time-averaged RMS,
which together capture both the loudness ceiling and the mean energy.

```{r env-loop}
env_results <- vector("list", nrow(syl_windows))

for (i in seq_len(nrow(syl_windows))) {
  row <- syl_windows[i, , drop = FALSE]

  # amp_env() returns a numeric vector of envelope values
  env <- amp_env(
    row,
    wav_dir       = wav_dir,
    msmooth       = c(256, 50),
    amp_normalize = "peak",
    plot          = FALSE
  )

  env_results[[i]] <- data.frame(
    motif_index = i,
    start_time  = row$start_time,
    peak_amp    = max(env,             na.rm = TRUE),
    mean_amp    = mean(env,            na.rm = TRUE),
    rms_amp     = sqrt(mean(env^2,     na.rm = TRUE))
  )
}

env_df <- do.call(rbind, env_results)
knitr::kable(env_df, digits = 4)
```

---

## Phase 4: Combine and Visualize Feature Trajectories

Merge all three feature tables and look at how each metric changes across
consecutive motif renditions.

```{r combine-features}
feature_df <- Reduce(
  function(a, b) merge(a, b, by = c("motif_index", "start_time")),
  list(entropy_df, ff_df, env_df)
)

knitr::kable(feature_df, digits = 4)
```

### Average entropy across renditions

```{r plot-entropy, fig.height=4}
plot(
  feature_df$motif_index,
  feature_df$mean_entropy,
  type  = "b",
  pch   = 16,
  col   = "steelblue",
  xlab  = "Motif index",
  ylab  = "Mean Wiener entropy",
  main  = "Spectral entropy across consecutive motif renditions",
  xaxt  = "n"
)
axis(1, at = feature_df$motif_index)

# ±1 SD ribbon
arrows(
  x0     = feature_df$motif_index,
  y0     = feature_df$mean_entropy - feature_df$sd_entropy,
  y1     = feature_df$mean_entropy + feature_df$sd_entropy,
  length = 0.04,
  angle  = 90,
  code   = 3,
  col    = "steelblue"
)
```

### Mean fundamental frequency across renditions

```{r plot-ff, fig.height=4}
plot(
  feature_df$motif_index,
  feature_df$mean_ff_kHz,
  type  = "b",
  pch   = 16,
  col   = "tomato",
  xlab  = "Motif index",
  ylab  = "Mean F0 (kHz)",
  main  = "Fundamental frequency across consecutive motif renditions",
  xaxt  = "n"
)
axis(1, at = feature_df$motif_index)

arrows(
  x0     = feature_df$motif_index,
  y0     = feature_df$mean_ff_kHz - feature_df$sd_ff_kHz,
  y1     = feature_df$mean_ff_kHz + feature_df$sd_ff_kHz,
  length = 0.04,
  angle  = 90,
  code   = 3,
  col    = "tomato"
)
```

### Envelope amplitude across renditions

```{r plot-env, fig.height=4}
plot(
  feature_df$motif_index,
  feature_df$mean_amp,
  type  = "b",
  pch   = 16,
  col   = "darkorange",
  ylim  = range(c(feature_df$mean_amp, feature_df$peak_amp), na.rm = TRUE),
  xlab  = "Motif index",
  ylab  = "Amplitude (peak-normalized)",
  main  = "Amplitude envelope across consecutive motif renditions",
  xaxt  = "n"
)
axis(1, at = feature_df$motif_index)

lines(feature_df$motif_index, feature_df$peak_amp,
      type = "b", pch = 1, col = "darkorange", lty = 2)

legend("topright",
  legend = c("Mean amplitude", "Peak amplitude"),
  col    = "darkorange",
  pch    = c(16, 1),
  lty    = c(1, 2),
  bty    = "n"
)
```

---

## Exporting the Feature Table

The combined `feature_df` is a standard R data frame ready for any downstream
analysis — linear models, mixed effects models, or further visualization.

```{r export-features, eval=FALSE}
# Save as CSV for use in other tools or scripts
write.csv(feature_df,
  file      = "motif_features.csv",
  row.names = FALSE
)
```

---

## Summary

This vignette demonstrated a complete single-recording pipeline that bridges
motif detection and acoustic feature analysis:

| Phase | Key steps | Output |
|---|---|---|
| Motif detection | `detect_template()` → `find_motif()` | `motifs` data frame |
| Syllable window | Fixed offset from `start_time` | `syl_windows` data frame |
| Entropy | `spectral_entropy()` per row | Mean + SD per motif |
| Pitch | `FF()` per row, voiced frames only | Mean F0 ± SD per motif |
| Envelope | `amp_env()` per row | Peak, mean, RMS per motif |
| Visualization | Base R `plot()` | Trajectory plots across renditions |

The same pattern — iterate per row, collect scalar summaries, bind into a
data frame — scales directly to larger datasets by replacing the `for` loop
with `lapply()` or `parallel_apply()`, or by passing a multi-row data frame
directly to `spectral_entropy()` or `FF()`.

### Relationship to SAP generic functions

This workflow — detect motifs, build syllable windows, iterate per-row feature
extraction, collect scalar summaries — is exactly what the SAP-method generics
(`spectral_entropy.Sap()`, `FF.Sap()`, `plot_heatmap.Sap()`) do internally.
The difference is that the SAP methods wrap these steps into a single function
call and extend them to handle multiple WAV files spread across multiple
directories (one per day). Working through the single-file version here makes
it easier to understand what those functions are doing under the hood and to
interpret their outputs.

---

## Session Info

```{r session-info}
sessionInfo()
```
