---
title: "Advanced Acoustic Feature Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Acoustic Feature Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

This vignette shows how to detect motifs in a single recording and extract
quantitative acoustic features from a selected syllable across repeated
renditions.

**Prerequisites**: Before reading this vignette, we recommend completing:

- [Overview: ASAP 101](single_wav_analysis.html)
- [Motif Detection](motif_detection.html)
- [Acoustic Feature Analysis](acoustic_feature_analysis.html)

**What you will learn**:

1. How to recap motif detection in a few key lines
2. How to define a syllable window of interest relative to motif boundaries
3. How to iterate `spectral_entropy()`, `FF()`, and `amp_env()` over all
   motifs to build a per-motif feature table
4. How to summarize and visualize feature trajectories across renditions

---

## Overview.

The workflow has three steps:

1. Detect repeated motifs in the recording
2. Define a consistent syllable window within each motif
3. Measure entropy, pitch, and amplitude for that window and summarize the
   results across renditions

---

## Setup

```{r setup}
library(ASAP)

wav_file <- system.file("extdata", "zf_example.wav", package = "ASAP")
wav_dir <- dirname(wav_file)
```

---

## Motif Detection

The visualization and parameter-tuning steps are covered in detail in the
[Motif Detection](motif_detection.html) vignette. Here we recap only the
core lines needed to produce the `motifs` table that drives Phase 2.

```{r motif-detection, fig.keep='none'}
# 1. Extract a reference clip and build a template from syllable "d"
clip_path <- create_audio_clip(wav_file, start_time = 1, end_time = 2.5)

template <- create_template(clip_path,
  template_name  = "d",
  start_time     = 0.72,
  end_time       = 0.84,
  freq_min       = 1,
  freq_max       = 10,
  write_template = FALSE
)

# 2. Detect template occurrences across the full recording
template_matches <- detect_template(
  x                = wav_file,
  template         = template,
  proximity_window = 1,
  save_plot        = FALSE
)

# 3. Expand each detection into full-motif boundaries
motifs <- find_motif(template_matches,
  pre_time = 0.7,
  lag_time = 0.5,
  wav_dir  = wav_dir
)

knitr::kable(motifs, digits = 3)
```

---

## Selecting a Syllable Window of Interest

Rather than analyzing the entire motif, we focus on a specific syllable to
reduce noise and make comparisons more precise. We define the syllable window
as a fixed offset **relative to each motif's `start_time`**.

Based on visual inspection of the motif spectrogram (see the
[Motif Detection](motif_detection.html) vignette), syllable "e" falls
roughly 0.78–0.85 s after the motif onset. We encode these offsets once so
they are applied consistently across every motif row.

```{r syllable-window}
# Offsets from each motif's start_time (seconds) — syllable "e"
syl_pre <- 0.8 # syllable onset relative to motif start
syl_post <- 0.85 # syllable offset relative to motif start

# Build a per-motif data frame for the syllable window
syl_windows <- motifs
syl_windows$start_time <- motifs$start_time + syl_pre
syl_windows$end_time <- motifs$start_time + syl_post
syl_windows$duration <- syl_post - syl_pre

knitr::kable(syl_windows[, c("filename", "start_time", "end_time", "duration")],
  digits = 3
)
```

Use `visualize_segments()` to inspect the syllable "e" window across **all**
detected motifs at once. Consistent spectrogram structure across panels
confirms the offset is correctly aligned before committing it to the
feature-extraction loop.

```{r visualize-syllable-windows, fig.height=4, fig.width=1.5}
visualize_segments(syl_windows,
  wav_dir   = wav_dir,
  n_samples = nrow(syl_windows)
)
```

## Feature Extraction Across All Motifs

We now iterate over every row of `syl_windows` and compute three features.
Each function accepts a single-row data frame plus `wav_dir`, so the pattern
is uniform for all three.

### 3a. Spectral Entropy

```{r entropy-loop}
entropy_results <- vector("list", nrow(syl_windows))

for (i in seq_len(nrow(syl_windows))) {
  row <- syl_windows[i, , drop = FALSE]

  # spectral_entropy() returns a 2-column matrix: [time, entropy]
  ent_mat <- spectral_entropy(
    row,
    wav_dir    = wav_dir,
    method     = "wiener",
    normalize  = FALSE,
    plot       = FALSE
  )

  n_frames <- sum(!is.na(ent_mat[, 2]))
  entropy_results[[i]] <- data.frame(
    motif_index  = i,
    start_time   = row$start_time,
    mean_entropy = mean(ent_mat[, 2], na.rm = TRUE),
    sem_entropy  = sd(ent_mat[, 2], na.rm = TRUE) / sqrt(n_frames)
  )
}

entropy_df <- do.call(rbind, entropy_results)
knitr::kable(entropy_df, digits = 4)
```

### 3b. Fundamental Frequency

```{r ff-loop}
ff_results <- vector("list", nrow(syl_windows))

for (i in seq_len(nrow(syl_windows))) {
  row <- syl_windows[i, , drop = FALSE]

  # FF() returns a 2-column matrix: [time, frequency (kHz)]
  ff_mat <- FF(
    row,
    wav_dir   = wav_dir,
    method    = "cepstrum",
    fmax      = 1000,
    threshold = 10,
    plot      = FALSE
  )

  # Exclude zeros (unvoiced frames)
  voiced <- ff_mat[ff_mat[, 2] > 0, 2]

  ff_results[[i]] <- data.frame(
    motif_index = i,
    start_time  = row$start_time,
    mean_ff_kHz = ifelse(length(voiced) > 0, mean(voiced, na.rm = TRUE), NA_real_),
    sem_ff_kHz  = ifelse(length(voiced) > 0, sd(voiced, na.rm = TRUE) / sqrt(length(voiced)), NA_real_)
  )
}

ff_df <- do.call(rbind, ff_results)
knitr::kable(ff_df, digits = 4)
```

### 3c. Amplitude Envelope

We summarize the envelope by its peak amplitude and the time-averaged RMS,
which together capture both the loudness ceiling and the mean energy.

```{r env-loop}
env_results <- vector("list", nrow(syl_windows))

for (i in seq_len(nrow(syl_windows))) {
  row <- syl_windows[i, , drop = FALSE]

  # amp_env() returns a numeric vector of envelope values
  env <- amp_env(
    row,
    wav_dir       = wav_dir,
    msmooth       = c(256, 50),
    plot          = FALSE
  )

  n_env <- sum(!is.na(env))
  env_results[[i]] <- data.frame(
    motif_index = i,
    start_time  = row$start_time,
    mean_amp    = mean(env, na.rm = TRUE),
    sem_amp     = sd(env, na.rm = TRUE) / sqrt(n_env),
    rms_amp     = sqrt(mean(env^2, na.rm = TRUE))
  )
}

env_df <- do.call(rbind, env_results)
knitr::kable(env_df, digits = 4)
```

---

## Combine and Visualize Feature Trajectories

Merge all three feature tables and look at how each metric changes across
consecutive motif renditions.

```{r combine-features}
feature_df <- Reduce(
  function(a, b) merge(a, b, by = c("motif_index", "start_time")),
  list(entropy_df, ff_df, env_df)
)

knitr::kable(feature_df, digits = 4)
```

### Average entropy across renditions

```{r plot-entropy, fig.height=4, fig.width=4}
plot(
  feature_df$motif_index,
  feature_df$mean_entropy,
  type  = "b",
  pch   = 16,
  col   = "steelblue",
  ylim  = range(c(feature_df$mean_entropy - feature_df$sem_entropy,
                  feature_df$mean_entropy + feature_df$sem_entropy), na.rm = TRUE),
  xlab  = "Motif index",
  ylab  = "Mean Wiener entropy",
  main  = "Spectral entropy across consecutive motif renditions",
  xaxt  = "n"
)
axis(1, at = feature_df$motif_index)

# ±1 SEM ribbon
arrows(
  x0     = feature_df$motif_index,
  y0     = feature_df$mean_entropy - feature_df$sem_entropy,
  y1     = feature_df$mean_entropy + feature_df$sem_entropy,
  length = 0.04,
  angle  = 90,
  code   = 3,
  col    = "steelblue"
)
```

### Mean fundamental frequency across renditions

```{r plot-ff, fig.height=4, fig.width=4}
plot(
  feature_df$motif_index,
  feature_df$mean_ff_kHz,
  type  = "b",
  pch   = 16,
  col   = "tomato",
  ylim  = range(c(feature_df$mean_ff_kHz - feature_df$sem_ff_kHz,
                  feature_df$mean_ff_kHz + feature_df$sem_ff_kHz), na.rm = TRUE),
  xlab  = "Motif index",
  ylab  = "Mean F0 (kHz)",
  main  = "Fundamental frequency across consecutive motif renditions",
  xaxt  = "n"
)
axis(1, at = feature_df$motif_index)

arrows(
  x0     = feature_df$motif_index,
  y0     = feature_df$mean_ff_kHz - feature_df$sem_ff_kHz,
  y1     = feature_df$mean_ff_kHz + feature_df$sem_ff_kHz,
  length = 0.04,
  angle  = 90,
  code   = 3,
  col    = "tomato"
)
```

### Envelope amplitude across renditions

```{r plot-env, fig.height=4, fig.width=4}
plot(
  feature_df$motif_index,
  feature_df$mean_amp,
  type  = "b",
  pch   = 16,
  col   = "darkorange",
  ylim  = range(c(feature_df$mean_amp - feature_df$sem_amp,
                  feature_df$mean_amp + feature_df$sem_amp), na.rm = TRUE),
  xlab  = "Motif index",
  ylab  = "Mean Amplitude",
  main  = "Amplitude envelope across consecutive motif renditions",
  xaxt  = "n"
)
axis(1, at = feature_df$motif_index)

arrows(
  x0     = feature_df$motif_index,
  y0     = feature_df$mean_amp - feature_df$sem_amp,
  y1     = feature_df$mean_amp + feature_df$sem_amp,
  length = 0.04,
  angle  = 90,
  code   = 3,
  col    = "darkorange"
)
```

---

## Exporting the Feature Table

The combined `feature_df` is a standard R data frame ready for any downstream
analysis.

```{r export-features, eval=FALSE}
# Save as CSV for use in other tools or scripts
write.csv(feature_df,
  file      = "syllable_features.csv",
  row.names = FALSE
)
```

---

## Summary

This vignette demonstrated a complete single-recording pipeline that bridges
motif detection and acoustic feature analysis:

| Phase | Key steps | Output |
|---|---|---|
| Motif detection | `detect_template()` → `find_motif()` | `motifs` data frame |
| Syllable window | Fixed offset from `start_time` | `syl_windows` data frame |
| Entropy | `spectral_entropy()` per row | Mean ± SEM per motif |
| Pitch | `FF()` per row, voiced frames only | Mean F0 ± SEM per motif |
| Envelope | `amp_env()` per row | Mean ± SEM, RMS per motif |
| Visualization | Base R `plot()` | Trajectory plots across renditions |

The same pattern — iterate per row, collect scalar summaries, bind into a
data frame — scales directly to larger datasets by replacing the `for` loop
with `lapply()` or `parallel_apply()`, or by passing a multi-row data frame
directly to `spectral_entropy()` or `FF()`.

---

## Session Info

```{r session-info}
sessionInfo()
```
