---
title: "Finding Motifs in Zebra Finch Song"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Finding Motifs in Zebra Finch Song}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

This vignette demonstrates how to find and extract **song motifs** from a single 
zebra finch WAV file using ASAP's template matching workflow. A motif is the 
stereotyped sequence of syllables that adult male zebra finches repeat during 
singing.

**Purpose of this vignette**: This tutorial provides the fundamental concepts for 
template-based motif detection in ASAP. You'll learn how to interactively 
optimize template parameters using a single recording before applying them to 
bulk processing. The emphasis is on developing a reliable "anchor" template 
that will serve as the foundation for detecting motifs across longitudinal 
recordings.

The workflow involves:

1. **Visualizing the song** - Identify a clear motif rendition
2. **Creating an audio clip** - Extract a segment containing the motif
3. **Creating a template** - Select a distinctive syllable as an anchor point
4. **Detecting the template** - Find all occurrences of the template
5. **Extracting motifs** - Define motif boundaries around template detections
6. **Visualizing results** - View extracted motifs and amplitude heatmaps

> **Iterative Optimization**: Steps 3-6 should be repeated multiple times 
> to refine template parameters until reliable detection is achieved. This 
> optimization is crucial before proceeding to bulk processing.

## Setup

```{r setup}
library(ASAP)

# Get path to example WAV file included with the package
wav_file <- system.file("extdata", "zf_example.wav", package = "ASAP")
```

## 1. Visualize the Song

First, let's look at the full recording to identify motif structure:

```{r visualize-full, fig.height=4}
visualize_song(wav_file)
```

We can see multiple motif renditions in this recording. Let's zoom in on a 
clear section to identify a good reference motif:

```{r visualize-segment, fig.height=4}
visualize_song(wav_file, 
               start_time_in_second = 1, 
               end_time_in_second = 2.5)
```

## 2. Create an Audio Clip

To create a template, we first extract an audio clip containing a clear motif.
We'll use the segment from 1-2.5 seconds which contains a complete motif.

```{r create-clip}
# Create an audio clip from the WAV file
clip_path <- create_audio_clip(wav_file,
                               start_time = 1,
                               end_time = 2.5)

# View the created clip path
clip_path
```

## 3. Create a Template (Critical Step)

Creating a good template is **the most critical step** in the motif detection 
workflow. The template serves as an "anchor point" that ASAP uses to locate 
motifs throughout recordings. A poorly chosen template will result in missed 
detections or false positives.

### Characteristics of a Good Template Syllable

When selecting a syllable for template creation, look for one that:

- **Is acoustically distinctive** - Has unique spectral features that 
  differentiate it from other syllables and background noise
- **Occurs exactly once per motif** - Ensures one-to-one mapping between 
  detections and motifs
- **Has clear temporal boundaries** - Sharp onset and offset for precise 
  alignment
- **Is consistently produced** - Appears reliably across all motif renditions
- **Is stable across development** - If analyzing longitudinal data, choose 
  a syllable that remains recognizable over time

### Template Selection Process

Based on the spectrogram above, we'll use syllable "d" which occurs around 
0.72-0.84 seconds within our clip. Let's visualize this specific region:

```{r create-template, fig.height=4, fig.width=3, out.width="40%"}
# First, visualize the template region
visualize_song(wav_file,
               start_time_in_second = 1 + 0.72,
               end_time_in_second = 1 + 0.84)
```

Now create the template with specified frequency bounds:

```{r make-template, out.width="40%"}
# Create the template
template <- create_template(clip_path,
                            template_name = "d",
                            start_time = 0.72,
                            end_time = 0.84,
                            freq_min = 1,
                            freq_max = 10,
                            write_template = FALSE)

# View template info
template
```

### Key Template Parameters

| Parameter | Description | Optimization Tips |
|-----------|-------------|-------------------|
| `start_time/end_time` | Syllable boundaries | Adjust to capture complete syllable without gaps |
| `freq_min/freq_max` | Frequency range (kHz) | Narrow range reduces noise; 1-10 kHz typical for zebra finch |

## 4. Detect Template Occurrences

Now we search for all occurrences of this template throughout the recording:

```{r detect-template, fig.height=6}
# Run template detection on the original WAV file
# proximity_window filters out multiple detections within the same motif
template_matches <- detect_template(x = wav_file,
                                    template = template,
                                    proximity_window = 1,  # Filter detections within 1s
                                    save_plot = FALSE)

# View detection results
knitr::kable(template_matches, digits = 2)
```

### Key Detection Parameters

| Parameter | Description | Purpose |
|-----------|-------------|--------|
| `proximity_window` | Time window (seconds) to filter nearby detections | Eliminates false positives within a motif duration. Only the highest-scoring detection within each window is retained. |
| `threshold` | Minimum correlation score (0-1) | Set during template creation; detections below this score are discarded(available for sap method) |

### Evaluating Detection Quality

Review the detection results carefully:

- **Score column**: Higher scores indicate better matches. If scores are 
  uniformly low, the template may need adjustment.
- **Number of detections**: Should match expected motif count based on 
  visual inspection.
- **Detection spacing**: Detections should be evenly spaced if motifs occur 
  regularly.

If detection results are unsatisfactory, **return to Step 3** and try:

- Different syllable selection
- Adjusting time boundaries
- Modifying frequency range

## 5. Extract Motif Boundaries

Once we have reliable template detections, we define motif boundaries by 
extending a fixed time window before and after each detection:

```{r extract-motifs}
# Define motif boundaries around each detection
# pre_time: how much before the template to include
# lag_time: how much after the template to include
motifs <- find_motif(template_matches,
                     pre_time = 0.7,
                     lag_time = 0.5,
                     wav_dir = dirname(wav_file))

# View extracted motifs
knitr::kable(motifs, digits = 2)
```

### Adjusting Pre/Lag Times

| Parameter | Description | Adjustment Strategy |
|-----------|-------------|---------------------|
| `pre_time` | Time before template anchor | Should capture syllables preceding the anchor |
| `lag_time` | Time after template anchor | Should capture syllables following the anchor |

## 6. Visualize Detected Motifs

Let's visualize the detected motifs to verify our detection worked correctly:

```{r visualize-motifs, fig.height=8}
# Visualize all extracted motifs
visualize_segments(motifs, 
                   wav_dir = dirname(wav_file),
                   n_samples = min(nrow(motifs), 4))
```

### Quality Check

Examine the extracted motifs:

- Are all motifs complete (no truncation at boundaries)?
- Are motifs properly aligned?
- Are there any false positives (non-motif sounds)?
- Is the temporal structure consistent across renditions?

If issues are found, **iterate back through Steps 3-6** with adjusted parameters.

## 7. Amplitude Envelope Heatmap

Once satisfied with detection quality, create a heatmap of amplitude envelopes 
across all detected motifs to visualize the temporal structure:

```{r plot-heatmap, fig.height=3, fig.width=8}
# Plot amplitude envelope heatmap
plot_heatmap(motifs, wav_dir = dirname(wav_file))
```

A well-detected set of motifs will show consistent vertical banding patterns 
corresponding to syllables across all renditions.

## Summary

This vignette demonstrated the core ASAP workflow for finding motifs in a single 
recording. The key insight is that **template optimization is an iterative process** 
— Steps 3-6 should be repeated until detection results are satisfactory.

| Step | Function | Description |
|------|----------|-------------|
| 1 | `visualize_song()` | View spectrogram to identify motifs |
| 2 | `create_audio_clip()` | Extract a reference motif segment |
| 3 | `create_template()` | Create a template from a distinctive syllable |
| 4 | `detect_template()` | Find all template occurrences |
| 5 | `find_motif()` | Define motif boundaries around detections |
| 6 | `visualize_segments()` | View extracted motif spectrograms |
| 7 | `plot_heatmap()` | Visualize amplitude envelope patterns |

### Key Parameters Reference

| Parameter | Description | Typical Value |
|-----------|-------------|---------------|
| `start_time/end_time` | Template time limits (seconds) | 0.1-0.2s duration |
| `freq_min/freq_max` | Frequency range (kHz) | 1-10 kHz for zebra finch |
| `pre_time` | Time before template for motif start | 0.2-0.8s |
| `lag_time` | Time after template for motif end | 0.2-0.8s |

## Next Steps: Bulk Processing with SAP Objects

Once you have optimized template parameters using a single recording 
(as demonstrated in this vignette), you can apply them to bulk processing 
of longitudinal recordings using SAP objects.

The following vignettes cover bulk processing workflows:

- **Constructing SAP Objects** - How to organize and import longitudinal 
  recording data
- **Detecting Motifs from Longitudinal Recordings** - Applying optimized 
  templates across multiple recordings

Here's a preview of the SAP object pipeline using the parameters we 
optimized above:

```{r full-pipeline, eval=FALSE}
# Create SAP object from organized recording folders
sap <- create_sap_object(
  base_path = "/path/to/recordings",
  subfolders_to_include = c("190", "201", "203"),
  labels = c("BL", "Post", "Rec")
)

# Run motif detection pipeline with optimized parameters
sap <- sap |>
  create_audio_clip(indices = 1, start_time = 1, end_time = 2.5, 
                    clip_names = "m1") |>
  create_template(template_name = "d", clip_name = "m1",
                  start_time = 0.72, end_time = 0.84,  # Optimized from this vignette
                  freq_min = 1, freq_max = 10, 
                  threshold = 0.5,                     # Filter low-scoring matches
                  write_template = TRUE) |>
  detect_template(template_name = "d",
                  threshold = 0.5,                     # Can adjust threshold here too
                  proximity_window = 1) |>             # Remove duplicate detections
  find_motif(template_name = "d", pre_time = 0.7, lag_time = 0.5) |>
  analyze_spectral(balanced = TRUE) |>
  find_clusters() |>
  run_umap()

# Visualize results
sap |>
  plot_heatmap(balanced = TRUE) |>
  plot_umap(split.by = "label")
```

### Bulk Processing Parameters

When processing longitudinal recordings with SAP objects, these additional 
parameters become important:

| Parameter | Location | Description |
|-----------|----------|-------------|
| `threshold` | `create_template()` / `detect_template()` | Minimum correlation score (0-1). Lower values = more detections (potentially more false positives). Can be adjusted in `detect_template()` to refine without recreating template. |
| `proximity_window` | `detect_template()` | Time window (seconds) for filtering duplicate detections. Set to ~motif duration to ensure one detection per motif. |

## Session Info

```{r session-info}
sessionInfo()
```
