---
title: "Motif Detection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Motif Detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

This vignette demonstrates how to find and extract **song motifs** from a single
zebra finch WAV file using ASAP's template matching workflow. A motif is the
stereotyped sequence of syllables that adult male zebra finches repeat during
singing.

This tutorial introduces the core logic of template-based motif detection in
ASAP. The emphasis is on interactively optimizing a reliable "anchor" template
using one recording before carrying the same strategy into longitudinal
analysis.

**Prerequisites**: Before reading this vignette, we recommend completing:

- [Overview: ASAP 101](single_wav_analysis.html) - Basic ASAP functions

**What you will learn**:

1. How to choose a good reference motif from a single recording
2. How to create and test a syllable template for motif detection
3. How to verify extracted motifs with spectrograms and heatmaps

---

## Overview

Motif detection in ASAP is an iterative single-file workflow: identify a clear
motif rendition, choose a distinctive anchor syllable, detect that template
throughout the recording, and then inspect the extracted motifs. Unlike the SAP
object tutorials, the goal here is not to present a full batch-processing
pipeline. Instead, this vignette focuses on the small decisions that make a
template robust before you scale up.

> **Iterative optimization**: Steps 3-6 should usually be repeated more than
> once. Refining the template is the most important part of getting reliable
> motif boundaries later in longitudinal analysis.

---

## Setup

```{r setup}
library(ASAP)

# Get path to example WAV file included with the package
wav_file <- system.file("extdata", "zf_example.wav", package = "ASAP")
```

---

## 1. Visualize the Song

First, let's look at the full recording to identify motif structure:

```{r visualize-full, fig.height=4}
visualize_song(wav_file)
```

We can see multiple motif renditions in this recording. Let's zoom in on a 
clear section to identify a good reference motif:

```{r visualize-segment, fig.height=4}
visualize_song(wav_file,
  start_time_in_second = 1,
  end_time_in_second = 2.5
)
```

This zoomed view helps you choose a clean motif rendition with clear syllable
boundaries and minimal overlap from neighboring vocalizations.

---

## 2. Create an Audio Clip

To create a template, we first extract an audio clip containing a clear motif.
We'll use the segment from 1-2.5 seconds which contains a complete motif.

```{r create-clip}
# Create an audio clip from the WAV file
clip_path <- create_audio_clip(wav_file,
  start_time = 1,
  end_time = 2.5
)

# View the created clip path
clip_path
```

---

## 3. Create a Template (Critical Step)

Creating a good template is **the most critical step** in the motif detection
workflow. The template serves as an "anchor point" that ASAP uses to locate
motifs throughout recordings. A poorly chosen template will result in missed
detections or false positives.

### Characteristics of a Good Template Syllable

When selecting a syllable for template creation, look for one that:

- **Is acoustically distinctive** - Has unique spectral features that 
  differentiate it from other syllables and background noise
- **Occurs exactly once per motif** - Ensures one-to-one mapping between 
  detections and motifs
- **Has clear temporal boundaries** - Sharp onset and offset for precise 
  alignment
- **Is consistently produced** - Appears reliably across all motif renditions
- **Is stable across development** - If analyzing longitudinal data, choose 
  a syllable that remains recognizable over time

### Template Selection Process

Based on the spectrogram above, we'll use syllable "d" which occurs around 
0.72-0.84 seconds within our clip. Let's visualize this specific region:

```{r create-template, fig.height=4, fig.width=3, out.width="40%", fig.align = "center"}
# First, visualize the template region
visualize_song(wav_file,
  start_time_in_second = 1 + 0.72,
  end_time_in_second = 1 + 0.84
)
```

Now create the template with specified frequency bounds:

```{r make-template, out.width="40%", fig.align = "center"}
# Create the template
template <- create_template(clip_path,
  template_name = "d",
  start_time = 0.72,
  end_time = 0.84,
  freq_min = 1,
  freq_max = 10,
  write_template = FALSE
)

# View template info
template
```

### Key Template Parameters

| Parameter | Description | Optimization Tips |
|-----------|-------------|-------------------|
| `start_time/end_time` | Syllable boundaries | Adjust to capture complete syllable without gaps |
| `freq_min/freq_max` | Frequency range (kHz) | Narrow range reduces noise; 1-10 kHz typical for zebra finch |

### Tuning tips

- If the template matches too many unrelated sounds, narrow `freq_min/freq_max`
  around the most distinctive part of the syllable.
- If good motif renditions are being missed, expand the time window slightly so
  the full syllable is captured.
- Prefer a syllable that appears once per motif; repeated syllables are more
  likely to produce ambiguous detections.

---

## 4. Detect Template Occurrences

Now we search for all occurrences of this template throughout the recording:

```{r detect-template, fig.height=6}
# Run template detection on the original WAV file
# proximity_window filters out multiple detections within the same motif
template_matches <- detect_template(
  x = wav_file,
  template = template,
  proximity_window = 1, # Filter detections within 1s
  save_plot = FALSE
)

# View detection results
knitr::kable(template_matches, digits = 2)
```

### Key Detection Parameters

| Parameter | Description | Purpose |
|-----------|-------------|--------|
| `proximity_window` | Time window (seconds) to filter nearby detections | Eliminates false positives within a motif duration. Only the highest-scoring detection within each window is retained. |
| `threshold` | Minimum correlation score (0-1) | Set during template creation; detections below this score are discarded(available for sap method) |

### Evaluating Detection Quality

Review the detection results carefully:

- **Score column**: Higher scores indicate better matches. If scores are 
  uniformly low, the template may need adjustment.
- **Number of detections**: Should match expected motif count based on 
  visual inspection.
- **Detection spacing**: Detections should be evenly spaced if motifs occur 
  regularly.

If detection results are unsatisfactory, **return to Step 3** and try:

- Different syllable selection
- Adjusting time boundaries
- Modifying frequency range

---

## 5. Extract Motif Boundaries

Once we have reliable template detections, we define motif boundaries by 
extending a fixed time window before and after each detection:

```{r extract-motifs}
# Define motif boundaries around each detection
# pre_time: how much before the template to include
# lag_time: how much after the template to include
motifs <- find_motif(template_matches,
  pre_time = 0.7,
  lag_time = 0.5,
  wav_dir = dirname(wav_file)
)

# View extracted motifs
knitr::kable(motifs, digits = 2)
```

### Adjusting Pre/Lag Times

| Parameter | Description | Adjustment Strategy |
|-----------|-------------|---------------------|
| `pre_time` | Time before template anchor | Should capture syllables preceding the anchor |
| `lag_time` | Time after template anchor | Should capture syllables following the anchor |

---

## 6. Visualize Detected Motifs

Let's visualize the detected motifs to verify our detection worked correctly:

```{r visualize-motifs, fig.height=8}
# Visualize all extracted motifs
visualize_segments(motifs,
  wav_dir = dirname(wav_file),
  n_samples = min(nrow(motifs), 4)
)
```

### Quality Check

Examine the extracted motifs:

- Are all motifs complete (no truncation at boundaries)?
- Are motifs properly aligned?
- Are there any false positives (non-motif sounds)?
- Is the temporal structure consistent across renditions?

If issues are found, **iterate back through Steps 3-6** with adjusted parameters.

---

## 7. Amplitude Envelope Heatmap

Once satisfied with detection quality, create a heatmap of amplitude envelopes 
across all detected motifs to visualize the temporal structure:

```{r plot-heatmap, fig.height=3, fig.width=8}
# Plot amplitude envelope heatmap
plot_heatmap(motifs, wav_dir = dirname(wav_file))
```

A well-detected set of motifs will show consistent vertical banding patterns 
corresponding to syllables across all renditions.

---

## Summary

This vignette demonstrated the core ASAP workflow for finding motifs in a single
recording. The key insight is that **template optimization is an iterative process**
— Steps 3-6 should be repeated until detection results are satisfactory.

| Step | Function | Description |
|------|----------|-------------|
| 1 | `visualize_song()` | View spectrogram to identify motifs |
| 2 | `create_audio_clip()` | Extract a reference motif segment |
| 3 | `create_template()` | Create a template from a distinctive syllable |
| 4 | `detect_template()` | Find all template occurrences |
| 5 | `find_motif()` | Define motif boundaries around detections |
| 6 | `visualize_segments()` | View extracted motif spectrograms |
| 7 | `plot_heatmap()` | Visualize amplitude envelope patterns |

### Key Parameters Reference

| Parameter | Description | Typical Value |
|-----------|-------------|---------------|
| `start_time/end_time` | Template time limits (seconds) | 0.1-0.2s duration |
| `freq_min/freq_max` | Frequency range (kHz) | 1-10 kHz for zebra finch |
| `pre_time` | Time before template for motif start | 0.2-0.8s |
| `lag_time` | Time after template for motif end | 0.2-0.8s |

## Next Steps: Longitudinal Recording Analysis with SAP Object

Once you have optimized template parameters using a single recording
(as demonstrated in this vignette), you can apply them to bulk processing
of longitudinal recordings using SAP objects.

The following vignettes cover the longitudinal analysis workflow:

- [**Constructing SAP Object**](construct_sap_object.html) - How to organize 
  and import longitudinal recording data
- [**Longitudinal Motif Detection**](longitudinal_motif_detection.html) - 
  Applying optimized templates across multiple recordings

---

## Session Info

```{r session-info}
sessionInfo()
```
