---
title: "Overview: ASAP 101"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Overview: ASAP 101}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

This vignette is the quickest way to get oriented with ASAP on a single WAV
file. While ASAP is designed for large-scale longitudinal studies, the core
functions work directly on individual recordings, which makes them a useful
starting point for learning the workflow.

The goal here is not to exhaust every function in the package. Instead, this
article introduces the basic sequence most users need when first opening a new
recording: inspect the song, detect bouts, segment syllables, and export a
clean example clip.

**What you will learn**:

1. How to inspect a WAV file with ASAP spectrogram tools
2. How to detect bouts and segment syllables in a single recording
3. How to export a detected bout as a standalone WAV clip

---

## Overview

Single-file analysis is the fastest way to understand how ASAP behaves before
moving on to SAP-object workflows. The same core functions used here also power
the larger longitudinal tutorials, but running them on one example file makes
it much easier to tune parameters and visually check the results.

## Setup

```{r setup}
library(ASAP)

# Get path to example WAV file included with the package
wav_file <- system.file("extdata", "zf_example.wav", package = "ASAP")
```

---

## 1. Audio Visualization

The `visualize_song()` function creates spectrogram visualizations of audio
recordings. This is often the first step when inspecting a new recording.

### Full recording

```{r visualize-full, fig.height=4}
visualize_song(wav_file)
```

### Specific time window

You can focus on a smaller time range with `start_time_in_second` and
`end_time_in_second`:

```{r visualize-segment, fig.height=4}
visualize_song(
  wav_file,
  start_time_in_second = 1,
  end_time_in_second = 4
)
```

Use this zoomed view to sanity-check whether the recording contains clear song,
where dense singing occurs, and which time window is a good candidate for the
next analysis steps.

---

## 2. Bout Detection

A "bout" is a continuous period of singing. The `find_bout()` function detects
bout boundaries using RMS amplitude thresholding after bandpass filtering.

```{r find-bout}
bouts <- find_bout(
  wav_file,
  rms_threshold = 0.1,
  min_duration = 0.7,
  plot = TRUE
)

knitr::kable(bouts, digits = 3)
```

### Key parameters

- `rms_threshold`: Higher values require louder sounds; lower values detect
  quieter vocalizations
- `min_duration`: Minimum bout length (filters out short sounds or noise)
- `freq_range`: Bandpass filter range (default: 1-8 kHz for zebra finch)

### Quality control

The overlaid detection plot is the main QC step for bout finding. Check whether
the detected start and end times match the visible song energy in the
spectrogram. If bouts are being split too aggressively, try increasing
`gap_duration` or lowering `rms_threshold`. If short noise events are being
mistaken for song, increase `min_duration`.

---

## 3. Syllable Segmentation

The `segment()` function detects individual syllables within a chosen time
window using dynamic spectral thresholding.

```{r segment, fig.height=6}
syllables <- segment(
  wav_file,
  start_time = 1,
  end_time = 5,
  flim = c(1, 8),
  silence_threshold = 0.01,
  min_syllable_ms = 20,
  max_syllable_ms = 240,
  min_level_db = 10,
  verbose = FALSE
)

knitr::kable(syllables, digits = 2)
```

### Understanding the output

The returned data frame contains:

- `start_time`/`end_time`: Syllable boundaries in seconds
- `duration`: Syllable length
- `silence_gap`: Gap before the next syllable
- `selec`: Selection number for tracking

### Tuning tips

| Parameter | Role | Typical effect |
|-----------|------|----------------|
| `silence_threshold` | Minimum silent gap between syllables | Lower values merge nearby events; higher values split more aggressively |
| `min_syllable_ms` | Short-event filter | Increase to suppress clicks and noise |
| `max_syllable_ms` | Long-event filter | Decrease to avoid merging long phrases into one syllable |
| `min_level_db` | Detection threshold | Raise it for noisier files; lower it for quieter syllables |

---

## 4. Export Bout Clips

Once you are happy with bout detection, you can export selected bouts as
standalone audio clips. Here we write the first detected bout to a temporary
directory, visualize the exported clip, and then remove the temporary files at
the end of the vignette. This keeps the example self-contained and avoids
leaving extra files behind after the vignette runs.

```{r export-bout}
bout_export_dir <- file.path(tempdir(), "asap_bout_export")
dir.create(bout_export_dir, recursive = TRUE, showWarnings = FALSE)
bout_export_meta <- NULL
exported_bout_file <- NULL

if (!is.null(bouts) && nrow(bouts) >= 1) {
  bout_export_meta <- create_bout_clips(
    bouts[1, , drop = FALSE],
    wav_dir = dirname(wav_file),
    output_dir = bout_export_dir,
    output_format = "wav",
    write_metadata = FALSE,
    verbose = FALSE
  )

  exported_bout_file <- bout_export_meta$output_path[1]

  knitr::kable(
    bout_export_meta[, c("clip_id", "start_time", "end_time", "duration", "output_path")],
    digits = 3
  )
}
```

```{r visualize-exported-bout, fig.height=4}
if (!is.null(exported_bout_file)) {
  visualize_song(exported_bout_file)
}
```

```{r cleanup-exported-bout, include = FALSE}
if (exists("bout_export_dir") && dir.exists(bout_export_dir)) {
  unlink(bout_export_dir, recursive = TRUE, force = TRUE)
}
```

The exported metadata table includes the clip name, original time boundaries,
duration, and the path to the generated file. In longer workflows this metadata
is useful for tracing exported clips back to the source recording.

---

## Next Steps

After you are comfortable with these basics, the next two single-recording
vignettes extend the workflow in two different directions:

- [Motif Detection](motif_detection.html) - Template matching on a single WAV file
- [Acoustic Feature Analysis](acoustic_feature_analysis.html) - Spectral entropy,
  pitch, and amplitude-envelope analysis

## Session Info

```{r session-info}
sessionInfo()
```
